Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-21 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify statistical modeling, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the apparent
disconnect between CJ's trait measurement and hypothesis testing
processes. Although separating these processes simplifies the analysis
of CJ data, it may also undermine the reliability of various statistical
results derived from the measurement.

To address these issues, the study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. As a result, the study divides its content into six
main sections. Section~\ref{sec-thurstone_theory} provides an overview
of Thurstone's theory. Section~\ref{sec-theory-issues} discusses the
identified issues in detail. Section~\ref{sec-theoretical} extends
Thurstone's general form to address these challenges. The extension
integrates core theoretical principles alongside key assessment design
features relevant to CJ experiments, such as the selection of judges,
stimuli, and comparisons. Section~\ref{sec-statistical} translates these
theoretical and practical elements into a probabilistic statistical
model to analyze dichotomous pairwise comparison data.
Section~\ref{sec-discussion} discusses the findings, limitations, and
challenges and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's conclusions.

@@

Despite its limitations, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, as
\citet[pp.~16]{vanderLinden_et_al_2017_I} noted, this substitution has
minimal impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation. Furthermore, this limited impact is supported by the
fact that the Normal and Logistic distributions exhibit analogous
statistical properties, differing only by a scaling factor of
approximately \(1.7\).

@@

Building on the previous section, it is clear that, despite its
limitations, researchers typically rely on the BTL model to measure a
trait and place the compared stimuli along such trait continuum
\citetext{\citealp[pp.~269]{Thurstone_1927b}; \citealp{Everitt_et_al_2010}}.
Additionally, the CJ literature shows that researchers frequently use
point estimates of BTL scores or their transformations to conduct
further analyses or hypothesis tests. For example, researchers have used
these scores to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty can bias the analysis and reduce
the precision of hypothesis tests. Notably, the direction and magnitude
of such biases are often unpredictable. Results may be attenuated,
exaggerated, or remain unaffected depending on the degree of uncertainty
in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues, showing how
neglecting the uncertainty associated with BTL scores can reduce the
precision of an effect linked to the predictor.

@@

A motivating example that will appear throughout the document clarifies
these concepts. In this example, researchers aim to determine: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, researcher
design a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges then
compare pairs of students' written texts (stimuli), to produce a
dichotomous outcome that reflects the relative quality of each text
(trait). Based on this setup, researchers reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Finally, following current CJ practices, researchers
then rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to a
\emph{census} of comparative data, that is, data representing all
repeated judgments made by every available judge for each pair of
stimuli produced by each pair of individuals in the population, hence
its name. Conversely, Section~\ref{sec-theory-theoretical_SC} integrates
the assessment design features into what we call the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

\subsubsection{Integrating initial theoretical
principles}\label{sec-theory-theoretical_P1}

@@

SCM~\ref{fig-cj03_scm} presents the initial integration of theoretical
principles into the conceptual-population model for the example, applied
to examine the effect of different teaching methods on students' writing
ability. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
variables. The subscripts \(h\) and \(i\) label the students who
authored the texts (individuals), while the indices \(a\) and \(b\)
represent the compared texts (stimuli). Moreover, the index \(j\) refers
to the judge, and \(k\) specifies the judgment number, accounting for
experimental conditions where judges perform the same judgment multiple
times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system allows
comparisons between texts written by the same student (\(h = i\),
\(a \neq b\)) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself (\(h = i\), \(a = b\)),
as such comparison lacks practical meaning within the CJ framework.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments developed under the first scenario
of Section~\ref{sec-theory-issue1b}, and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
the judges' biases \((B_{kj})\). Together with the outcome, these
variables form the preliminary set of endogenous variables,
\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence, while the absence of an edge implies no relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally represent observed variables
with solid black circles and unobserved (latent) variables with open
circles \citep{Morgan_et_al_2014}. They also display conditioned
variables by surrounding them with a square. Here, \emph{conditioning}
refers to restricting the statistical analysis to a subset of the
population defined by the values of the given variable
\citep[pp.~32]{Neal_2020}. While DAGs often omit exogenous variables for
simplicity, following a \emph{standard representation}, this section
adopts the \emph{magnified representation} of DAGs, as including
exogenous variables can improve clarity and help reveal potential issues
related to conditioning and confounding \citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the causal relationships among
variables based on Thurstone's theory. Particularly, the graph shows how
the texts' discriminal processes \((T_{ha}, T_{ib})\) influence their
discriminal difference \((D_{hiabjk})\), which in turn determines the
outcome \((O^{cp}_{hiabjk})\). It also depicts the judges' biases
\((B_{kj})\) as an additional factor influencing the texts' discriminal
difference. Furthermore, the DAG distinguish between observed endogenous
variables, such as the outcome, and latent endogenous variables, which
include the texts' discriminal processes, their discriminal difference,
and the judges' biases.

\subsubsection{The comparative census data
structure}\label{sec-theory-theoretical_P2}

Although specifying a system's data structure is not mandatory when
using SCMs and DAGs, defining one can improve clarity and facilitate the
description of the system. Thus, to re-express the scalar form of the CJ
system, outlined in the previous section, into an simpler equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:

@@

where an element of \(I\) corresponds to a unique individual \(h\) or
\(i\), while each row of \(IA\) corresponds to a unique pairing of
individuals \(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents
the number of individuals, while \(n_{A}\) denotes the number of stimuli
available per individual. Thus, vector \(I\) has a length of \(n_{I}\),
and matrix \(IA\) has a dimension of \((n_{I} \cdot n_{A})\) rows and
\(2\) columns. The vector and the matrix consist of scalars from the set
of natural numbers \(\mathbb{N}\).

Similarly, an element of the vector \(J\) corresponds to a unique judge
\(j\). Moreover, each row of \(JK\) specifies a unique pairing of a
judge \(j\) with a judgment number \(k\). In this case, \(n_{J}\)
indicates the total number of judges, and \(n_{K}\) represents the
number of judgments each judge performs. As a result, vector \(J\) has a
length of \(n_{J}\), and matrix \(JK\) has a dimension of
\((n_{J} \cdot n_{K})\) rows and \(2\) columns. Like the previous
structures, the vector and the matrix consist of scalars from the set of
natural numbers \(\mathbb{N}\).

Additionally, we define the matrix \(RPUV\), where each row uniquely
pairs the corresponding rows from the matrices \(IA\) and \(JK\). This
matrix reflects the fundamental idea that the discriminal difference of
the stimuli becomes an observable outcome only through the judges'
perceptions. Thus, the matrix is defined as follows:

@@

This matrix also consists of scalars from the set of natural numbers
\(\mathbb{N}\) and has a dimension of
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
\(6\) columns. The matrix's dimension results from the
conceptual-population model's assumption that researchers have access to
a \emph{census} of comparative data, capturing all repeated judgments
made by every available judge for each pair of stimuli produced by each
pair of individuals in the population. Here, the term
\({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial coefficient,
which quantifies the total number of individual-stimuli comparison pairs
within the conceptual population. The coefficient is defined as follows:

\[
{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
\]

We visualize the structure of these vectors and matrices by considering
small values for the parameters corresponding to number of individuals,
the number of available stimuli per individual, the total number of
judges, and the number of judgments each judge performs. For instance,
assuming \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\),
the matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
take the following form:

@@

Thus, using the data structure defined by the matrices in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}), we can re-express
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in a simpler
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
judges' biases \(B_{JK}\) are represented as vectors rather than scalar
values. These vectors encompass all values from the conceptual
population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
length \({n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
\(D_{RPUV}\) is a latent vector of the same dimension. Additionally,
\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model incorporates two \emph{hierarchical
structural components}. Each component describes how different observed
or latent variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition addresses the problems arising
when an additional hierarchical structure is relevant to the stimuli, as
outlined in the second scenario of Section~\ref{sec-theory-issue1b}.
Furthermore, it facilitates analyses that integrate both trait
measurement and hypothesis testing processes, thereby accounting for
both traits and their associated uncertainties, reducing bias and
improving measurement precision (see Section~\ref{sec-theory-issue2}).

Figure~\ref{fig-cj09} displays the two hierarchical structural
components for the example, applied to examine the effect of different
teaching methods on students' writing ability. Specifically, the top
branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
component describes how the students' writing-quality trait \((T_{I})\)
and a set of \emph{relevant} text-related variables \((X_{IA})\), such
as text length, causally influence the texts' written-quality trait
\((T_{IA})\). The DAG also highlights the idiosyncratic errors
\((e_{IA})\), which capture variations in the texts' trait that neither
the students' trait nor other relevant text-related variables can
explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors of length
\(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively. Additionally,
\(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns.

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issues}. These enhancements include
accounting for judges' biases and integrating hierarchical structural
components. Nevertheless, despite its promise of enhancing measurement
accuracy and precision, the model still depends on the unrealistic
assumption that researchers have access to a \emph{census} of
comparative data, encompassing all repeated judgments from every judge
for each pair of stimuli produced by every individual in the population.
Since researchers rarely meet this assumption in practice, they must
consider a more realistic scenario. The extensions to the
conceptual-population model, derived from this realistic scenario, are
concisely described in the sample-comparison model.

@@

Following the convention of \citet[pp.~499-516]{McElreath_2020}, the
DAG~\ref{fig-cj12_dag} depicts the conceptual-population outcome
\(O^{cp}_{RPUV}\) as an unobserved variable, emphasizing the
researchers' inability to access such an outcome due to the sampling
mechanism. Moreover, the DAG shows the conditioned \emph{sample design}
variable \(S\), which causally influence the sample-comparison outcome
\(O^{sc}_{RPUV}\). This variable reflects how this sample-comparison
outcome is a sample drawn from the conceptual-population outcome
\(O^{cp}_{RPUV}\). Specifically, the variable is a vector
\(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\) corresponds to the number
of rows in the matrix \(RPUV\) (refer to
Section~\ref{sec-theory-theoretical_P2}), and each \(s_{i}\) is a binary
variable indicating the presence \((1)\) or absence \((0)\) of data in
the corresponding positions of the conceptual-population outcome
\(O^{cp}_{RPUV}\). Specifically,

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify statistical modeling, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the disconnect
between CJ's trait measurement and hypothesis testing processes.
Although separating these processes simplifies the analysis of CJ data,
it may also undermine the reliability of various statistical results
derived from the measurement.

To address these issues, this study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
BTL model assumptions to actively testing whether those assumptions fit
the data under analysis.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key assessment design features relevant
to CJ experiments, such as the selection of judges, stimuli, and
comparisons. Section~\ref{sec-statistical} translates these theoretical
and practical elements into a probabilistic statistical model to analyze
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the findings, limitations, and challenges and explores avenues
for future research. Finally, Section~\ref{sec-conclusion} summarizes
the study's conclusions.

@@

Despite its limitations, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, this substitution has minimal
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation
\citetext{\citealp[pp.~16]{vanderLinden_et_al_2017_I}; \citealp{McElreath_2022}}.
Furthermore, this limited impact is supported by the fact that the
Normal and Logistic distributions exhibit analogous statistical
properties, differing only by a scaling factor of approximately \(1.7\).

@@

Building on the previous section, it is clear that, despite its
limitations, researchers typically rely on the BTL model to measure a
trait and place the compared stimuli along its continuum
\citep[pp.~269]{Thurstone_1927b}. Additionally, the CJ literature shows
that researchers frequently use point estimates of BTL scores or their
transformations to conduct further analyses or hypothesis tests. For
example, researchers have used these scores to identify `misfit' judges
and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other assessment methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the underlying trait of interest
\citep{Casalicchio_et_al_2015, Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

Nevertheless, while separating the trait measurement and hypothesis
testing processes simplifies the analysis of CJ data, the statistical
literature cautions against relying solely on the point estimates of BTL
scores to conduct further analyses or hypothesis tests, as this practice
can undermine the resulting statistical conclusions. A key consideration
is that BTL scores are parameter estimates that inherently carry
uncertainty (measurement error). Ignoring this uncertainty can bias the
analysis and reduce the precision of hypothesis tests. The direction and
magnitude of such biases are often unpredictable. Results may be
attenuated, exaggerated, or remain unaffected depending on the degree of
uncertainty in the scores and the actual effects being tested
\citetext{\citealp[pp.~25]{Kline_et_al_2023}; \citealp[pp.~137]{Hoyle_et_al_2023}}.
Furthermore, the reduced precision in hypothesis tests diminishes their
statistical power, increasing the likelihood of committing type-I or
type-II errors \citep{McElreath_2020}.
Figure~\ref{fig-measurement_error} illustrates these issues, showing how
neglecting measurement error in an outcome can reduce the precision of a
predictor' estimated effect.

@@

A motivating example that will appear throughout the document clarifies
these concepts. In this example, researchers aim to determine: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, a researcher
designs a CJ experiment by randomly assigning students (individuals) to
two groups, each receiving a different teaching method. Judges then
compare pairs of students' written texts (stimuli) to produce a
dichotomous outcome reflecting the relative quality of each text
(trait). Based on this setup, researchers can reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers then rely on
estimates from the BTL model, or its transformations, to approximate
this estimand.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to Causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we refer to as the
\emph{conceptual-population model}. This model assumes that researchers
have access to a ``conceptual population'' of comparative data, that is,
data representing all repeated judgments made by every available judge
for each pair of stimuli produced by each pair of individuals in the
population. Conversely, Section~\ref{sec-theory-theoretical_SC}
integrates the assessment design features into what we refer to as the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence. The absence of an edge implies no direct relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. They also display conditioned variables by
surrounding them with a square. Here, \emph{conditioning} refers to
restricting the statistical analysis to a subset of the population
defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
Although DAGs often omit exogenous variables for simplicity, this
section includes exogenous variables to improve clarity and reveal
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

\subsubsection{Integrating the first theoretical
principles}\label{sec-theory-theoretical_P1}

SCM~\ref{fig-cj03_scm} illustrates the first theoretical principles of
the conceptual-population model designed to examine the impact of
different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{hiabjk})\) and several related variables. The subscripts
\(h\) and \(i\) label the students who authored the texts (individuals),
while the indices \(a\) and \(b\) represent the compared texts
(stimuli). The index \(j\) refers to the judge, and \(k\) represents the
judgment index, accounting for experimental conditions where judges
perform the same judgment multiple times, i.e., a \emph{repeated
measures design} \citep[pp.~366-376]{Lawson_2015}. Notably, the indexing
system facilitates comparisons between texts written by the same student
(\(h = i\), \(a \neq b\)) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself (\(h = i\), \(a = b\)),
as such comparison lacks practical relevance within the CJ framework.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
difference \((D_{hiabjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments developed in
Section~\ref{sec-theory-issue1b} and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
the judges' biases \((B_{kj})\). Together with the outcome, these
variables constitute the preliminary set of endogenous variables,
\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

Moreover, Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the expected causal outlined in
Thurstone's theory. The graph shows that the discriminal processes of
the texts \((T_{ha}, T_{ib})\) influence their discriminal difference
\((D_{hiabjk})\), which in turn determines the outcome
\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
differentiates between observed endogenous variables, such as the
outcome, and latent endogenous variables, including the texts'
discriminal processes, their discriminal difference, and the judges'
biases.

\subsubsection{The conceptual-population data
structure}\label{sec-theory-theoretical_P2}

Although specifying a system's data structure is not mandatory when
using SCMs and DAGs, defining one can improve clarity and facilitate the
description of the system. Thus, to re-express the scalar form of the CJ
system outlined in the previous section into an equivalent vectorized
form, we first define the vectors \(I\) and \(J\), along with the
matrices \(IA\) and \(JK\), as follows:

@@

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\((n_{I} \cdot n_{A})\) rows and \(2\) columns, where \(n_{A}\)
specifies the number of stimuli available per individual. Likewise, each
row of \(JK\) associates a judge \(j\) with a judgment index \(k\),
where \(n_{K}\) indicates the number of judgments each judge makes.
Consequently, the matrix \(JK\) has \((n_{J} \cdot n_{K})\) rows and
\(2\) columns.

Additionally, we define the matrix \(RPUV\) that assigns each row from
the \(IA\) matrix to a corresponding row from the \(JK\) matrix. Thus,
the matrix consists of
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
\(6\) columns. Here, the term \({n_{I} \cdot n_{A} \choose 2}\)
represents the Binomial coefficient, which quantifies the total number
of repeated judgments each judge makes for every pair of stimuli
generated by each pair of individuals in the population. The matrix is
defined as follows:

@@

It is easier to visualize the structure of these vectors and matrices by
considering an example where \(n_{I} = 5\), \(n_{A} = 2\),
\(n_{J} = 3\), and \(n_{K} = 3\). In this simple case, the vectors and
matrices described in equations (\ref{eq-mat12}) and (\ref{eq-mat3})
take the following form:

@@

Then, using equations (\ref{eq-mat12}) and (\ref{eq-mat3}), we can
re-express SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an
equivalent vectorized form, as shown in Figure~\ref{fig-cj04}. In this
depiction, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
judges' biases \(B_{JK}\) are represented as vectors rather than scalar
values. These vectors capture all the observations from the conceptual
population. Specifically, \(O^{cp}_{RPUV}\) is an observed vector of
length \(n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), and
\(D_{RPUV}\) is a latent vector of the same dimension. Moreover,
\(T_{IA}\) and \(B_{JK}\) are latent vectors of lengths
\((n_{I} \cdot n_{A})\) and \((n_{J} \cdot n_{K})\), respectively.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model incorporates two \emph{hierarchical
structural components}. A structural component describes how different
observed or latent variables influence the primary latent variable of
interest \citep{Everitt_et_al_2010}. Moreover, their hierarchical nature
enable analyses when the stimuli exhibit a hierarchical structure (refer
to Section~\ref{sec-theory-issue1b}) and when researchers aim to test
hypotheses while accounting for the uncertainties associated with the
traits (see Section~\ref{sec-theory-issue2}).

Figure~\ref{fig-cj09} displays these components applied to examine the
effect of different teaching methods on students' writing ability. The
top branch of DAG~\ref{fig-cj09_dag} depicts the first component. This
component describes how the students' writing-quality trait \((T_{I})\)
and a set of \emph{relevant} text-related variables \((X_{IA})\), such
as text length, causally influence the texts' written-quality trait
\((T_{IA})\). The DAG also highlights the idiosyncratic errors
\((e_{IA})\), which capture variations in the texts' trait that neither
the students' trait nor other relevant text-related variables can
explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors of length
\(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively. Additionally,
\(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns.

@@

Overall, the conceptual-population model extends Thurstone's general
form by introducing key innovations to address the limitations discussed
in Section~\ref{sec-theory-issues}. These enhancements include
accounting for judges' biases and integrating hierarchical structural
components. Nevertheless, despite its promise of enhancing measurement
accuracy and precision, the model still depends on the unrealistic
assumption that researchers have access to a \emph{conceptual
population} of comparative data, encompassing all repeated judgments
from every judge for each pair of stimuli produced by every individual
in the population. Since researchers rarely meet this assumption in
practice, they must consider a more realistic scenario. The extensions
to the conceptual-population model, derived from this realistic
scenario, are concisely described in the sample-comparison model.

@@

Following the convention of \citet[pp.~499-516]{McElreath_2020} and
\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} depicts
the conceptual-population outcome \(O^{cp}_{RPUV}\) as an unobserved
variable, emphasizing the researchers' inability to access such an
outcome due to the sampling mechanism. Moreover, the DAG shows the
conditioned \emph{sample design} variable \(S\), which causally
influence the sample-comparison outcome \(O^{sc}_{RPUV}\) . This
variable reflects how this sample-comparison outcome is a sample drawn
from the conceptual-population outcome \(O^{cp}_{RPUV}\). Specifically,
the variable is a vector \(S = \{ s_{1}, \dots, s_{n} \}\), where \(n\)
corresponds to the number of rows in the matrix \(RPUV\) (refer to
Section~\ref{sec-theory-theoretical_P2}), and each \(s_{i}\) is a binary
variable indicating the presence \((1)\) or absence \((0)\) of data in
the corresponding positions of the conceptual-population outcome
\(O^{cp}_{RPUV}\). Specifically,
